Perihelion advance of Mercury

### 2020-10-07 perihelion_advance_of_mercury.Rmd
more work, mostly clarifying motivation for the methods used

and above we see that the orbit is accurately closed, as expected
(three orbits are simulated, the lines overlap).

@@

```{r,label=orbitsimulatorGRlongtime,cache=TRUE}
out <- trajectory(merc_perihelion, m=d, h=h, tau=seq(from=0,to=35e16,len=1e7))
```

We need to isolate the change points, here found isolating iterations
$i$ such that `r[i-1]<r[i]` and `r[i]>r[i+1]`:

@@

```{r,precessionwithclassicaldynamics,cache=TRUE}
out <- trajectory(merc_perihelion, m=d, h=h, tau=seq(from=0,to=35e16,len=1e7),GR=FALSE)
jj1 <- c(NA,out$rdot)
jj2 <- c(out$rdot,NA)
change <- (jj1>0) & (jj2<0)
tee    <- out$time[change]
phi <- out$phi[change]
phi <- (phi) %% (2*pi)
plot(tee,phi-pi,main='orbital angle at of Mercury at perihelion, classical case',xlab='time (seconds)',ylab='angle of perihelion')
```

>>>>

and above we see that the orbit is accurately closed, as expected
(three orbits are simulated, the lines overlap).  However, a more
exacting test would be to plot the same lines but focussing in on the
aphelion point $(30,0)$:

```{r,label=closeupofaphelion}
dx <- 0.01
dy <- 0.5
plot(jj$x,jj$y,xlim=c(29.99,30.001),ylim=c(-dy,dy),type='b',xlab='x',ylab='y')
```

In the above we see an accurately closed orbit (which simulates more
than three full revolutions round the sun).

@@

```{r,label=orbitsimulatorGRlongtime,cache=TRUE}
out <- trajectory(merc_perihelion,m=d,h=h,tau=seq(from=0,to=35e16,len=1e7)) # takes ~5 mins
tail(out)
```

From the above we see that the trajectory has been simulated up to
about $965$ radians or about $\frac{965}{2\pi}\simeq 153$ revolutions
round the sun.  Plotting the entire trajectory is computationally
expensive, but we can plot the simulated trajectory close to aphelion
as for the Newtonian case:

```{r, label-plotmercuryorbitGR}
maxx <- 46001200027
n <- 650000
s <- seq_len(n)
plot(out$x[s],out$y[s],xlim=c(maxx-500,maxx+100),ylim=c(-5e6,5e6),type='o',xlab="x",ylab="y",pch=c(16,rep(1,n-1)))
```

In the above plot, note the differing axis scales.  The start point is
shown as a solid circle, subseqent points shown with open circles.
Lines connect consecutive points in the simulation.  It is difficult
to see evidence for precession in this plot because the points
correspond to regular snapshots of Mercury's position in successive
orbits, and the time evolution is not shown.  To proceed we need to
find the aphelion (or an approximation to it) somehow, and show that
the aphelion systematically increases with time.  Given the numerics,
we find points on the simulated trajectory which are a local maximum
distance from the sun.  This is straightfoward numerically, we
identify iterations $i$ (which have radius `r[i]`) satisfying both
`r[i-1]<r[i]` and `r[i]>r[i+1]`:

@@

```{r,precessionwithclassicaldynamics,cache=TRUE}
out <- trajectory(merc_perihelion, m=d, h=h, tau=seq(from=0,to=35e16,len=1e7),GR=FALSE)
# trajectory() takes ~ 5 mins to run
jj1 <- c(NA,out$rdot)
jj2 <- c(out$rdot,NA)
change <- (jj1>0) & (jj2<0)
tee    <- out$time[change]
phi <- out$phi[change]
phi <- (phi) %% (2*pi)
plot(tee,phi-pi,main='orbital angle at of Mercury at perihelion, classical case',xlab='time (seconds)',ylab='angle of perihelion')
```
